---
title: "WSTC Timecourse Cytokine Analysis (MSD, LDH, TEER)"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{=html}
<!-- ============================================================
  Factors maintained throughout:
    CELLTYPE : SAE | LAE   (CELLTYPE_LEVELS)
    HORMONE  : NONE | Estradiol  (HORMONE_LEVELS)
    TIMEPOINT: 4 | 144   (TIMEPOINT_LEVELS)
    SEX      : M | F       (SEX_LEVELS)
    EXPOSURE : PBS Control | Pine 25 | Peat 25 | Eucalyptus 25 | Red Oak 25
  ============================================================ -->
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 8
)
set.seed(4454)
```

------------------------------------------------------------------------

# Setup

## Load Libraries & Configuration

```{r load-packages}
source(here::here("_load_all.R"))

suppressPackageStartupMessages({
  library(lme4)
  library(emmeans)
  library(variancePartition)
  library(cowplot)
  library(ggnewscale)
  library(ggpattern)
  library(ARTool)
})

cat("✓ All packages loaded\n")
cat("  Project root:", here::here(), "\n")
```

------------------------------------------------------------------------

# Data Loading & Preparation

## Load Raw Data

```{r load-data}
WSTC_highdose       <- readr::read_csv(here::here(PATH_DATA_RAW, "4hvs144h.csv"))
metadata_WSTC1_1152 <- readr::read_csv(here::here(PATH_DATA_RAW, "timecourse_metadata.csv"))

# LDH / TEER — uncomment when data is available
# LDH_TEER      <- readr::read_csv(here::here(PATH_DATA_RAW, "2023_LDHTEER_MJR1-191_4H.csv"))
# LDH_TEER$LDH  <- as.numeric(LDH_TEER$LDH)

cat("✓ Raw data loaded\n")
cat("  MSD rows:", nrow(WSTC_highdose),
    "| Metadata rows:", nrow(metadata_WSTC1_1152), "\n")
```

## Build WSTC_full

```{r build-wstc-full}
WSTC_avg  <- average_nonzero_by_sample(WSTC_highdose)
WSTC_full <- build_sala_full(WSTC_avg, metadata_WSTC1_1152)

assert_analysis_metadata(WSTC_full)

names(metadata_WSTC1_1152)

# temporary: do NOT filter until exposure mapping is confirmed
# WSTC_full <- WSTC_full %>% dplyr::filter(EXPOSURE != "Untreated_Control")

cat("✓ WSTC_full built\n")
cat("  Rows    :", nrow(WSTC_full), "\n")
cat("  CELLTYPE:", paste(levels(WSTC_full$CELLTYPE), collapse = " | "), "\n")
cat("  HORMONE :", paste(levels(WSTC_full$HORMONE),  collapse = " | "), "\n")
cat("  TIMEPOINT:", paste(levels(WSTC_full$TIMEPOINT), collapse = " | "), "\n")
cat("  SEX     :", paste(levels(WSTC_full$SEX),      collapse = " | "), "\n")
cat("  EXPOSURE:", paste(levels(WSTC_full$EXPOSURE), collapse = " | "), "\n")

cat("\nObserved EXPOSURE values:\n")
print(sort(unique(as.character(WSTC_full$EXPOSURE))))
cat("Missing EXPOSURE:", sum(is.na(WSTC_full$EXPOSURE)), "\n")
```

```{r define-plot-exposures}
plot_exposures <- intersect(
  ALL_HIGH_DOSE_EXPOSURES,
  unique(as.character(WSTC_full$EXPOSURE))
)

cat("Configured exposures:", paste(ALL_HIGH_DOSE_EXPOSURES, collapse = " | "), "\n")
cat("Present in WSTC_full:", paste(sort(unique(as.character(WSTC_full$EXPOSURE))), collapse = " | "), "\n")
cat("Using plot_exposures:", paste(plot_exposures, collapse = " | "), "\n")

if (length(plot_exposures) == 0) {
  stop("No configured high-dose exposures found in WSTC_full.")
}
```

------------------------------------------------------------------------

# LDH / TEER Analysis

```{r ldh-teer, eval=FALSE}
# Set eval=TRUE when LDH_TEER data is available

LDH_TEER[LDH_TEER == ""] <- NA
LDH_TEER_CLEAN <- na.omit(LDH_TEER)

LDH_TEER_AVG <- LDH_TEER_CLEAN %>%
  dplyr::group_by(PATIENTCODE, CELLTYPE, EXPOSURE, TIMEPOINT,
                  SEX, GROUP, CONCENTRATION) %>%
  dplyr::summarize(dplyr::across(c(LDH, TEERPRE, TEERPOST, TEERCHANGE), mean),
                   .groups = "drop")

LDH_Cytotox <- LDH_TEER_AVG %>%
  dplyr::group_by(PATIENTCODE) %>%
  dplyr::mutate(
    Cytotox = (LDH - LDH[GROUP == "Control"]) /
              (LDH[GROUP == "Positive Control"] - LDH[GROUP == "Control"]) * 100
  ) %>%
  dplyr::ungroup()

cat("Shapiro-Wilk — Cytotoxicity:\n")
print(shapiro.test(LDH_Cytotox$Cytotox))
cat("Shapiro-Wilk — TEER change:\n")
print(shapiro.test(LDH_Cytotox$TEERCHANGE))
```

------------------------------------------------------------------------


```{r paired-audit-and-enforce}
target_exposures <- intersect(
  plot_exposures,
  unique(as.character(WSTC_full$EXPOSURE))
)

# 1) Donor-level presence table
pair_audit <- WSTC_full %>%
  dplyr::filter(EXPOSURE %in% c(PBS_LEVEL, target_exposures)) %>%
  dplyr::distinct(PATIENTCODE, CELLTYPE, HORMONE, SEX, TIMEPOINT, EXPOSURE) %>%
  dplyr::mutate(present = 1L) %>%
  tidyr::pivot_wider(
    names_from = EXPOSURE, values_from = present, values_fill = 0
  )

print(pair_audit, n = 200)

# 2) For each exposure, keep only donor-strata with BOTH PBS and that exposure
enforce_pairs_for_exposure <- function(df, exposure, pbs = PBS_LEVEL) {
  diag <- log_paired_matching_results(
    df = df,
    exposure = exposure,
    pbs = pbs,
    verbose = TRUE
  )
  diag$paired
}

# 3) Build paired-only dataset across exposures
WSTC_paired <- dplyr::bind_rows(lapply(target_exposures, function(exp) {
  enforce_pairs_for_exposure(WSTC_full, exp)
})) %>%
  dplyr::distinct()

# 4) Verify strict 1:1 at donor-stratum level
pair_check <- WSTC_paired %>%
  dplyr::distinct(PATIENTCODE, CELLTYPE, HORMONE, SEX, TIMEPOINT, EXPOSURE) %>%
  dplyr::count(CELLTYPE, HORMONE, SEX, TIMEPOINT, EXPOSURE, name = "n")

print(pair_check, n = 200)

cat("✓ Paired dataset rows:", nrow(WSTC_paired), "\n")

pair_check_by_exposure <- purrr::map_dfr(plot_exposures, function(exp) {
  d <- enforce_pairs_for_exposure(WSTC_full, exp)

  chk <- d %>%
    dplyr::distinct(PATIENTCODE, CELLTYPE, HORMONE, SEX, TIMEPOINT, EXPOSURE) %>%
    dplyr::count(PATIENTCODE, CELLTYPE, HORMONE, SEX, TIMEPOINT, name = "n_levels") %>%
    dplyr::mutate(target_exposure = exp)

  bad <- chk %>% dplyr::filter(n_levels != 2)

  dplyr::tibble(
    target_exposure = exp,
    n_rows = nrow(d),
    n_strata = nrow(chk),
    n_bad_strata = nrow(bad)
  )
})

print(pair_check_by_exposure)

pair_check_counts <- WSTC_paired %>%
  dplyr::count(EXPOSURE, CELLTYPE, HORMONE, SEX, TIMEPOINT, name = "n_rows")

print(pair_check_counts, n = 200)
```

# Cytokine & Imputation Setup

Imputes values below LLOD (LOD/√2) using a single global pass over `WSTC_paired`.
The **sole cytokine-inclusion gate** is `impute_lod_sqrt2()`'s `ZERO_CUTOFF` (≤ 30% NA/zero
within each model stratum). The old dotplot-specific > 50% imputed filter is dropped: cytokine
validity is now determined consistently by the LMER pipeline (`valid_cytokines` inside
`run_lmer_chunk()`), which applies the identical `ZERO_CUTOFF`. The global `imp`/`df_use`
objects are used only for visualization (log₂FC, bar heights, raw-data points).

```{r cytokine-setup}
# Identify cytokine columns present in WSTC_paired once, outside the loop
cytokine_cols_all <- intersect(names(WSTC_paired), cytokine_llod$Analyte)

# Global imputation for visualization (dotplots, bar plots, smoker variance PCA).
# Statistical testing uses the per-stratum imputation inside run_lmer_chunk().
imp    <- impute_lod_sqrt2(
  input_data    = WSTC_paired,
  cols          = cytokine_cols_all,
  cytokine_llod = cytokine_llod,
  zero_cutoff   = ZERO_CUTOFF
)
df_use <- imp$data

cat("✓ Global imputation complete\n")
cat("  Valid cytokines:", length(imp$valid_cytokines), "\n")
cat("  Skipped (>", ZERO_CUTOFF * 100, "% zeros):",
    nrow(imp$skipped_cytokines), "\n\n")

cat("valid global cytokines length:", length(imp$valid_cytokines), "\n")
cat("First few global cytokines:", paste(head(imp$valid_cytokines), collapse = ", "), "\n")
```

--------

# LMER Significance Tables

Runs the 14-step lmer → significance-table pipeline for each of the four CELLTYPE × HORMONE strata. `run_lmer_chunk()` lives in `functions_analysis.R` and calls `exposure_lmer_pairwise()`, `interaction_lmer_pairwise()`, `summarize_to_wide()`, and `build_lmer_sig_table()`. LMER formulas include TIMEPOINT as a fixed effect when both `4` and `144` are present. Each call also returns `lmer_plot_data` — a tidy tibble (SEX / CELLTYPE / HORMONE / TIMEPOINT / EXPOSURE / CYTOKINE / q / sig) used by the dotplot section below as the single source of statistical truth.

## LAE — No Hormone

```{r lmer-lae-none}
LAE_NONE_res  <- run_lmer_chunk("LAE_NONE", "LAE", "NONE", sala_full = WSTC_paired)
LAE_NONE_sig  <- LAE_NONE_res$sig_table
LAE_NONE_lmer_interaction <- LAE_NONE_res$lmer_interaction
```

## LAE — Estradiol

```{r lmer-lae-e2}
LAE_E2_res  <- run_lmer_chunk("LAE_E2", "LAE", "Estradiol", sala_full = WSTC_paired)
LAE_E2_sig  <- LAE_E2_res$sig_table
LAE_E2_lmer_interaction <- LAE_E2_res$lmer_interaction
```

## SAE — No Hormone

```{r lmer-sae-none}
SAE_NONE_res  <- run_lmer_chunk("SAE_NONE", "SAE", "NONE", sala_full = WSTC_paired)
SAE_NONE_sig  <- SAE_NONE_res$sig_table
SAE_NONE_lmer_interaction <- SAE_NONE_res$lmer_interaction
```

## SAE — Estradiol

```{r lmer-sae-e2}
SAE_E2_res  <- run_lmer_chunk("SAE_E2", "SAE", "Estradiol", sala_full = WSTC_paired)
SAE_E2_sig  <- SAE_E2_res$sig_table
SAE_E2_lmer_interaction <- SAE_E2_res$lmer_interaction
```

## Combine LMER plot data

```{r lmer-combine}
# Merge per-stratum lmer_plot_data into one master tibble.
# This is the SINGLE source of q-values and significance calls for all plots.
lmer_results_all <- dplyr::bind_rows(
  LAE_NONE_res$lmer_plot_data,
  LAE_E2_res$lmer_plot_data,
  SAE_NONE_res$lmer_plot_data,
  SAE_E2_res$lmer_plot_data
)
cat("✓ lmer_results_all:", nrow(lmer_results_all), "rows\n")
cat("  Significant (q <", ALPHA_Q, "):", sum(lmer_results_all$sig, na.rm = TRUE), "\n")

stopifnot("TIMEPOINT" %in% names(lmer_results_all))

diag_tp <- lmer_results_all %>%
  dplyr::count(SEX, CELLTYPE, HORMONE, TIMEPOINT, EXPOSURE, name = "n_tests")

print(diag_tp)

stopifnot(all(c("4", "144") %in% unique(as.character(lmer_results_all$TIMEPOINT))))

# Long-format export: raw values + LMER significance per cytokine/exposure/sex/timepoint
WSTC_TMR_sig <- WSTC_paired %>%
  dplyr::filter(HORMONE == "NONE") %>%
  dplyr::mutate(TIMEPOINT = as.character(TIMEPOINT)) %>%
  dplyr::select(
    SAMPLEID, CELLTYPE, PATIENTCODE, TIMEPOINT, SEX, EXPOSURE, HORMONE,
    dplyr::any_of("SMOKER"),
    dplyr::any_of(cytokine_cols_all)
  ) %>%
  tidyr::pivot_longer(
    cols      = dplyr::any_of(cytokine_cols_all),
    names_to  = "CYTOKINE",
    values_to = "VALUE"
  ) %>%
  dplyr::left_join(
    lmer_results_all %>%
      dplyr::filter(HORMONE == "NONE") %>%
      dplyr::mutate(TIMEPOINT = as.character(TIMEPOINT)) %>%
      dplyr::select(SEX, CELLTYPE, HORMONE, TIMEPOINT, EXPOSURE, CYTOKINE, q, sig),
    by = c("SEX", "CELLTYPE", "HORMONE", "TIMEPOINT", "EXPOSURE", "CYTOKINE")
  ) %>%
  dplyr::mutate(
    sig = ifelse(is.na(sig), FALSE, sig)   # PBS rows / untested combos = not sig
  )

save_table(WSTC_TMR_sig, "WSTC_no_ESTRADIOL_with_sig.csv")
cat("✓ WSTC_no_ESTRADIOL_with_sig.csv saved (includes sig/q columns)\n")
```

## Timepoint Comparisons

Inference note: PBS is the exposure-control baseline **within each timepoint** (4h and 144h).
The primary timecourse inference is a difference-in-differences (DiD):
\[
[(\text{Exposure} - \text{PBS})_{144h}] - [(\text{Exposure} - \text{PBS})_{4h}]
\]
so conclusions are about change from 4h to 144h **relative to PBS**, not an unmeasured time baseline.

```{r timepoint-comparisons}

# Build a long table for timepoint modeling (from WSTC_paired wide cytokine matrix)
WSTC_full_long <- WSTC_paired %>%
  dplyr::mutate(TIMEPOINT = as.character(TIMEPOINT)) %>%
  dplyr::select(
    SAMPLEID, PATIENTCODE, CELLTYPE, HORMONE, TIMEPOINT, SEX, EXPOSURE,
    dplyr::any_of(cytokine_cols_all)
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::any_of(cytokine_cols_all),
    names_to = "CYTOKINE",
    values_to = "Value"
  )

# Keep only exposures present in this dataset
tp_exposures <- intersect(
  plot_exposures,
  unique(as.character(WSTC_full_long$EXPOSURE))
)

tp_stats <- WSTC_full_long %>%
  dplyr::filter(EXPOSURE %in% tp_exposures) %>%
  dplyr::group_by(CYTOKINE) %>%
  dplyr::group_modify(~ timepoint_lmer_within_exposure_sex(
    data = .x,
    response = "Value",
    min_donors = 4,
    time_levels = c("4", "144")
  )) %>%
  dplyr::ungroup()

save_table(tp_stats, "WSTC_timepoint_within_exposure_sex_lmer.csv")
cat("✓ Saved: WSTC_timepoint_within_exposure_sex_lmer.csv\n")
print(dplyr::arrange(tp_stats, p_adj_BH) %>% dplyr::slice_head(n = 20))

```

```{r timepoint-did-vs-pbs}
# Minimal input validation for DiD workflow
tp_levels_observed <- sort(unique(as.character(WSTC_full_long$TIMEPOINT)))
cat("Observed TIMEPOINT levels:", paste(tp_levels_observed, collapse = ", "), "\n")
stopifnot(all(c("4", "144") %in% tp_levels_observed))

did_stats <- build_timecourse_did_table(
  data = WSTC_full_long %>%
    dplyr::filter(EXPOSURE %in% c(PBS_LEVEL, tp_exposures)),
  response = "Value",
  ctrl_level = PBS_LEVEL,
  time_levels = c("4", "144"),
  min_donors = 4,
  pseudocount = PSEUDOCOUNT,
  alpha_q = ALPHA_Q,
  include_all_sex = TRUE
)

save_table(did_stats, "WSTC_timecourse_did_vs_pbs_lmer.csv")
cat("✓ Saved: WSTC_timecourse_did_vs_pbs_lmer.csv\n")

# Minimal output validation
if (nrow(did_stats) > 0) {
  tested_rows <- did_stats %>% dplyr::filter(!is.na(p.value))
  if (nrow(tested_rows) > 0) {
    stopifnot(all(!is.na(tested_rows$q.value)))
    stopifnot(all(is.finite(tested_rows$q.value)))
  }
}

print(
  did_stats %>%
    dplyr::arrange(q.value, dplyr::desc(abs(estimate))) %>%
    dplyr::select(
      cytokine, CELLTYPE, HORMONE, SEX, exposure, contrast,
      estimate, SE, conf.low, conf.high, p.value, q.value, direction, call
    ) %>%
    dplyr::slice_head(n = 30)
)
```

```{r timepoint-did-summary}
did_tested <- did_stats %>%
  dplyr::filter(!is.na(p.value), !is.na(q.value))

did_call_counts <- did_tested %>%
  dplyr::count(call, name = "n_tests") %>%
  dplyr::arrange(dplyr::desc(n_tests))

cat("DiD tested rows:", nrow(did_tested), "\n")
print(did_call_counts)

did_sig <- did_tested %>%
  dplyr::filter(q.value < ALPHA_Q) %>%
  dplyr::arrange(q.value, dplyr::desc(abs(estimate)))

cat("Significant DiD rows (q <", ALPHA_Q, "):", nrow(did_sig), "\n")

if (nrow(did_sig) > 0) {
  cat("Top significant DiD results (change from 4h to 144h relative to PBS control):\n")
  print(
    did_sig %>%
      dplyr::select(
        cytokine, CELLTYPE, HORMONE, SEX, exposure,
        estimate, conf.low, conf.high, p.value, q.value, direction, call
      ) %>%
      dplyr::slice_head(n = 15)
  )
} else {
  cat("No significant DiD rows at q <", ALPHA_Q,
      ": no significant exposure-associated change from 4h to 144h relative to PBS control.\n")
}
```

```{r timepoint-visualizations, fig.width=14, fig.height=8}
# choose top cytokines by BH-adjusted p-value from tp_stats
top_cyt <- tp_stats %>%
  dplyr::filter(!is.na(p_adj_BH), !is.na(CYTOKINE)) %>%
  dplyr::arrange(p_adj_BH) %>%
  dplyr::pull(CYTOKINE) %>%
  unique() %>%
  head(6)

cat("Top cytokines:", paste(top_cyt, collapse = ", "), "\n")
cat("WSTC_full_long rows:", nrow(WSTC_full_long), "\n")

for (cyt in top_cyt) {
  cat("Plotting:", cyt, "\n")
  p_sp <- plot_timepoint_spaghetti(
    df_long = WSTC_full_long,
    cytokine_name = cyt,
    y_log2 = TRUE
  )

  if (is.null(p_sp)) {
    cat("Skipping (no data):", cyt, "\n")
    next
  }

  print(p_sp)
  save_plot(
    paste0("timepoint_spaghetti_", safe_name(cyt), ".png"),
    p_sp, width = 14, height = 8, dpi = 300, bg = "white"
  )
}

p_hm <- plot_timepoint_delta_heatmap(
  WSTC_full_long %>%
    dplyr::filter(EXPOSURE %in% intersect(
      plot_exposures,
      unique(as.character(EXPOSURE))
    ))
)

if (!is.null(p_hm)) {
  print(p_hm)
  save_plot("timepoint_delta_heatmap.png", p_hm, width = 12, height = 9, dpi = 300, bg = "white")
} else {
  cat("Skipping heatmap (no data).\n")
}
```
--------

# Cytokine Dotplots

Loops over all 4 high-dose exposures in `plot_exposures`.\
For each exposure the pipeline:

1.  Filters `lmer_results_all` for the current exposure to get significance calls\
2.  Builds log₂FC relative to matched PBS (**visualization only** — not used for significance)\
3.  Produces a combined cowplot (pooled + sex-stratified dotplots) faceted by **CELLTYPE × HORMONE**\
4.  Saves per-cytokine bar plots and histograms, each faceted by **CELLTYPE × HORMONE**

```{r cytokine-dotplot, fig.width=14, fig.height=8, results='asis'}

# right before the failing check
cat("Observed EXPOSURE:\n")
print(sort(unique(as.character(WSTC_full$EXPOSURE))))
cat("Configured high-dose:\n")
print(plot_exposures)

# ── Shared color constants (match config.R) ───────────────────────────────────
UP_COLOR       <- "#D7191C"
DOWN_COLOR     <- "#2C7BB6"
ESTRADIOL_FILL <- "#E8F5E9"

# ── Per-exposure loop ─────────────────────────────────────────────────────────
available_high_dose_exposures <- intersect(
  plot_exposures,
  unique(as.character(WSTC_full$EXPOSURE))
)
missing_high_dose_exposures <- setdiff(plot_exposures, available_high_dose_exposures)

cat("Configured high-dose exposures found after normalization:",
    paste(available_high_dose_exposures, collapse = " | "), "\n")
if (length(missing_high_dose_exposures) > 0) {
  cat("Missing configured exposures after normalization:",
      paste(missing_high_dose_exposures, collapse = " | "), "\n")
}
if (length(available_high_dose_exposures) == 0) {
  stop("No configured high-dose exposures were found after normalization.")
}

for (target_exposure in available_high_dose_exposures) {

  exp_display     <- exposure_display_name(target_exposure)
  cat("\n\n## Exposure:", exp_display, "\n\n")

  # Derived labels used everywhere in this iteration
  exp_short       <- exposure_display_name(target_exposure, include_dose = FALSE)
  e2_label        <- paste(exp_short, "+ E2")
  hormone_levels  <- c(exp_short, e2_label)
  safe_exp        <- safe_name(target_exposure)
  border_color    <- unname(EXPOSURE_COLORS_DEEP[target_exposure])
  panel_fill_none <- unname(EXPOSURE_COLORS_LIGHT[target_exposure])
  if (!length(border_color) || is.na(border_color) ||
      !length(panel_fill_none) || is.na(panel_fill_none)) {
    cat("  \u2192 Missing configured colors after normalization; skipping plots.\n")
    next
  }

  # ── 1) Filter LMER results for this exposure ──────────────────────────────
  # lmer_results_all is the single source of q-values / sig calls (no separate
  # screen_one_exposure_lmer_log2() call; no >50% imputed pre-filter here).
  lmer_exp <- lmer_results_all %>%
    dplyr::filter(EXPOSURE == target_exposure)
  if (nrow(lmer_exp) == 0) {
    cat("  \u2192 No LMER rows found for this normalized exposure; skipping plots.\n")
    next
  }

  cat("  Cytokines tested by LMER:", length(unique(lmer_exp$CYTOKINE)), "\n")
  cat("  (Cytokines excluded by ZERO_CUTOFF are reported inside run_lmer_chunk)\n")

  # ── 2) Recode HORMONE display labels ────────────────────────────────────
  lmer_exp_F   <- lmer_exp %>% dplyr::filter(SEX == "F") %>%
    recode_hormone_labels(exp_short, e2_label, hormone_levels)
  lmer_exp_M   <- lmer_exp %>% dplyr::filter(SEX == "M") %>%
    recode_hormone_labels(exp_short, e2_label, hormone_levels)
  lmer_exp_All <- lmer_exp %>% dplyr::filter(SEX == "All") %>%
    recode_hormone_labels(exp_short, e2_label, hormone_levels)

  # ── 3) Determine significant cytokines from LMER ─────────────────────────
  keep_cytokines <- dplyr::bind_rows(lmer_exp_F, lmer_exp_M, lmer_exp_All) %>%
    dplyr::filter(sig) %>%
    dplyr::pull(CYTOKINE) %>%
    unique()

  cat("  Significant cytokines (q <", ALPHA_Q, "):", length(keep_cytokines), "\n")

  if (length(keep_cytokines) == 0) {
    cat("  → No significant cytokines; skipping plots.\n")
    next
  }

  # ── 4) log2FC relative to matched PBS (visualization only) ───────────────
  # log2FC drives dot size/position and bar height; it does NOT determine
  # significance, filter keep_cytokines, or influence q-values.
  long_target <- make_log2fc_long(df_use, keep_cytokines) %>%
    dplyr::filter(
      EXPOSURE == target_exposure,
      as.character(SEX) %in% c("F", "M")
    ) %>%
    dplyr::mutate(
      SEX       = factor(as.character(SEX), levels = SEX_LEVELS),
      TIMEPOINT = factor(as.character(TIMEPOINT), levels = TIMEPOINT_LEVELS)
    ) %>%
    recode_hormone_labels(exp_short, e2_label, hormone_levels)

  if (nrow(long_target) == 0) {
    cat("  No log2FC rows for this exposure.\n")
    next
  }

  # ── 5) Global y-axis order by mean |log2FC| (visualization only) ─────────
  cyt_order <- long_target %>%
    dplyr::group_by(CYTOKINE) %>%
    dplyr::summarise(score = mean(abs(log2FC), na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(score)) %>%
    dplyr::pull(CYTOKINE)

  # ── 6) Single-source significance maps (from LMER, not log2FC) ───────────
  sig_map_sex <- dplyr::bind_rows(lmer_exp_F, lmer_exp_M) %>%
  dplyr::filter(CYTOKINE %in% keep_cytokines) %>%
  dplyr::transmute(
    SEX       = factor(as.character(SEX), levels = SEX_LEVELS),
    CELLTYPE  = as.character(CELLTYPE),
    HORMONE   = factor(as.character(HORMONE), levels = hormone_levels),
    TIMEPOINT = factor(as.character(TIMEPOINT), levels = TIMEPOINT_LEVELS),
    CYTOKINE  = as.character(CYTOKINE),
    q         = as.numeric(q)
  ) %>%
  dplyr::group_by(SEX, CELLTYPE, HORMONE, TIMEPOINT, CYTOKINE) %>%
  dplyr::summarise(
    q = dplyr::if_else(all(is.na(q)), NA_real_, stats::median(q, na.rm = TRUE)),
    sig = !is.na(q) & q < ALPHA_Q,
    .groups = "drop"
  )

  sig_map_all <- lmer_exp_All %>%
  dplyr::filter(CYTOKINE %in% keep_cytokines) %>%
  dplyr::transmute(
    CELLTYPE  = as.character(CELLTYPE),
    HORMONE   = factor(as.character(HORMONE), levels = hormone_levels),
    TIMEPOINT = factor(as.character(TIMEPOINT), levels = TIMEPOINT_LEVELS),
    CYTOKINE  = as.character(CYTOKINE),
    q         = as.numeric(q)
  ) %>%
  dplyr::group_by(CELLTYPE, HORMONE, TIMEPOINT, CYTOKINE) %>%
  dplyr::summarise(
    q = dplyr::if_else(all(is.na(q)), NA_real_, stats::median(q, na.rm = TRUE)),
    sig_pooled = !is.na(q) & q < ALPHA_Q,
    .groups = "drop"
  )

 # ── 7) Build plot data frames ─────────────────────────────────────────────
make_plot_df <- function(df, grp_vars, sig_join, sig_col) {
  df %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(grp_vars))) %>%
    dplyr::summarise(mean_log2FC = mean(log2FC, na.rm = TRUE),
                     .groups = "drop") %>%
    dplyr::mutate(
      CYTOKINE = factor(as.character(CYTOKINE), levels = cyt_order)
    ) %>%
    add_rescaled_size("mean_log2FC", transform = "sqrt", out_col = "size_val") %>%
    dplyr::left_join(sig_join, by = grp_vars) %>%
    dplyr::mutate(
      q          = ifelse(is.na(q), 1, q),
      !!sig_col  := ifelse(is.na(.data[[sig_col]]), FALSE, .data[[sig_col]]),
      panel_fill = ifelse(as.character(HORMONE) == e2_label,
                          ESTRADIOL_FILL, panel_fill_none),
      direction  = dplyr::case_when(
        mean_log2FC > 0 ~ "Up",
        mean_log2FC < 0 ~ "Down",
        TRUE            ~ "Zero"
      )
    )
}

# Build FULL dataframes (all data, not filtered)
plot_df_pooled <- make_plot_df(
  long_target,
  c("CELLTYPE", "HORMONE", "TIMEPOINT", "CYTOKINE"),
  sig_map_all, "sig_pooled"
) %>%
  dplyr::mutate(All = factor("All", levels = "All"))

plot_df_sex <- make_plot_df(
  long_target %>%
    dplyr::mutate(SEX = factor(as.character(SEX), levels = SEX_LEVELS)),
  c("SEX", "CELLTYPE", "HORMONE", "TIMEPOINT", "CYTOKINE"),
  sig_map_sex, "sig"
) %>%
  dplyr::filter(as.character(SEX) %in% c("F", "M")) %>%
  dplyr::mutate(SEX = factor(as.character(SEX), levels = SEX_LEVELS))

cyt_levels <- levels(factor(plot_df_pooled$CYTOKINE, levels = cyt_order))

# ── 8) Filter to significant for DOTPLOTS only ─────────────────────────────
plot_df_pooled_sig <- plot_df_pooled %>%
  dplyr::filter(sig_pooled)

plot_df_sex_sig <- plot_df_sex %>%
  dplyr::filter(sig)

# ── 9) Build dotplots ──────────────────────────────────────────────────────
p_pooled <- make_dotplot(
  df           = plot_df_pooled_sig,
  x_var        = "All",
  title_str    = paste(exp_display, "\u2014 Pooled (All Samples)"),
  cyt_levels   = cyt_levels,
  border_color = border_color,
  up_color     = UP_COLOR,
  down_color   = DOWN_COLOR,
  show_y       = TRUE
)

p_sex <- make_dotplot(
  df           = plot_df_sex_sig,
  x_var        = "SEX",
  title_str    = paste(exp_display, "\u2014 By Sex"),
  cyt_levels   = cyt_levels,
  border_color = border_color,
  up_color     = UP_COLOR,
  down_color   = DOWN_COLOR,
  show_y       = FALSE
)

# ── 10) Combine with cowplot ──────────────────────────────────────────────
if (nrow(plot_df_pooled_sig) == 0 && nrow(plot_df_sex_sig) == 0) {
  cat("  \u2192 No significant plot rows remained after aggregation; skipping dotplots.\n")
  next
}

legend_source <- if (nrow(plot_df_pooled_sig) > 0) p_pooled else p_sex
legend_shared  <- cowplot::get_legend(legend_source)
p_pooled_noleg <- p_pooled + ggplot2::theme(legend.position = "none")
p_sex_noleg    <- p_sex    + ggplot2::theme(legend.position = "none")

p_combined <- cowplot::plot_grid(
  p_pooled_noleg, p_sex_noleg, legend_shared,
  nrow       = 1,
  rel_widths = c(1.2, 1.2, 0.4),
  align      = "h",
  axis       = "tb"
)

print(p_combined)

out_prefix <- paste0("dotplot_", safe_exp)
save_plot(paste0(out_prefix, "_combined.png"),
          p_combined, width = 14, height = 6, dpi = 300, bg = "white")
save_plot(paste0(out_prefix, "_combined_transparent.png"),
          p_combined, width = 14, height = 6, dpi = 300, bg = "transparent")
save_plot(paste0(out_prefix, "_pooled.png"),
          p_pooled,   width = 6,  height = 6, dpi = 300, bg = "white")
save_plot(paste0(out_prefix, "_sex.png"),
          p_sex,      width = 10, height = 6, dpi = 300, bg = "white")

  # ── 11) Per-cytokine bar plots & histograms ────────────────────────────────
  # make_cytokine_barplot() and make_cytokine_histogram() in functions_plots.R

  for (cyt in keep_cytokines) {

    donor_plate <- imp$data %>%
      dplyr::filter(EXPOSURE == target_exposure,
                    SEX %in% c("F", "M")) %>%
      dplyr::distinct(PATIENTCODE)

    d_raw <- dplyr::bind_rows(
      imp$data %>%
        dplyr::filter(SEX %in% c("F", "M"),
                      EXPOSURE == target_exposure) %>%
        dplyr::semi_join(donor_plate, by = "PATIENTCODE"),
      imp$data %>%
        dplyr::filter(SEX %in% c("F", "M"),
                      EXPOSURE == PBS_LEVEL) %>%
        dplyr::semi_join(donor_plate, by = "PATIENTCODE")
    ) %>%
      dplyr::mutate(
        HORMONE  = dplyr::recode(as.character(HORMONE),
                                 "NONE"      = exp_short,
                                 "Estradiol" = e2_label),
        HORMONE  = factor(HORMONE,  levels = hormone_levels),
        TIMEPOINT = factor(as.character(TIMEPOINT), levels = TIMEPOINT_LEVELS),
        SEX      = factor(as.character(SEX), levels = SEX_LEVELS),
        VALUE    = .data[[cyt]],
        EXPOSURE = factor(EXPOSURE, levels = c(PBS_LEVEL, target_exposure))
      )

    # Significance labels from single-source q map (LMER results only)
    sig_for_cyt <- sig_map_sex %>%
      dplyr::filter(CYTOKINE == cyt) %>%
      dplyr::select(CELLTYPE, HORMONE, TIMEPOINT, SEX, q, sig)

    male_pt_fill <- grDevices::adjustcolor(border_color, alpha.f = 1,
                                           red.f = 1.3, green.f = 1.3, blue.f = 1.3)

    summary_raw <- d_raw %>%
      dplyr::group_by(SEX, EXPOSURE, HORMONE, CELLTYPE, TIMEPOINT) %>%
      dplyr::summarise(
        mu = mean(VALUE, na.rm = TRUE),
        sd = sd(VALUE,   na.rm = TRUE),
        n  = dplyr::n(),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        panel_fill = ifelse(as.character(HORMONE) == e2_label,
                            ESTRADIOL_FILL, panel_fill_none),
        bar_fill   = border_color,
        bar_color  = border_color,
        direction  = dplyr::case_when(
          mu > 0 ~ "Up",
          mu < 0 ~ "Down",
          TRUE   ~ "Zero"
        ),
        sig_color  = dplyr::case_when(
          direction == "Up"   ~ UP_COLOR,
          direction == "Down" ~ DOWN_COLOR,
          TRUE                ~ "grey50"
        ),
        x_label = factor(
          dplyr::case_when(
            SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "M Control",
            SEX == "M" & EXPOSURE == target_exposure ~ paste("M", exp_short),
            SEX == "F" & EXPOSURE == PBS_LEVEL       ~ "F Control",
            SEX == "F" & EXPOSURE == target_exposure ~ paste("F", exp_short)
          ),
          levels = c("M Control", paste("M", exp_short),
                     "F Control", paste("F", exp_short))
        )
      ) %>%
      dplyr::left_join(sig_for_cyt, by = c("CELLTYPE", "HORMONE", "TIMEPOINT", "SEX")) %>%
      dplyr::mutate(
        q         = as.numeric(q),
        sig       = ifelse(is.na(sig), FALSE, sig),
        sig_label = dplyr::case_when(
          is.na(q)  ~ "",
          q < 0.001 ~ "***",
          q < 0.01  ~ "**",
          q < 0.05  ~ "*",
          TRUE      ~ ""
        )
      )

    d_raw <- d_raw %>%
      dplyr::mutate(
        x_label = factor(
          dplyr::case_when(
            SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "M Control",
            SEX == "M" & EXPOSURE == target_exposure ~ paste("M", exp_short),
            SEX == "F" & EXPOSURE == PBS_LEVEL       ~ "F Control",
            SEX == "F" & EXPOSURE == target_exposure ~ paste("F", exp_short)
          ),
          levels = c("M Control", paste("M", exp_short),
                     "F Control", paste("F", exp_short))
        ),
        point_fill  = dplyr::case_when(
          SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "grey80",
          SEX == "M" & EXPOSURE == target_exposure ~ male_pt_fill,
          TRUE                                     ~ "white"
        ),
        point_color = dplyr::case_when(
          EXPOSURE == PBS_LEVEL       ~ "grey40",
          EXPOSURE == target_exposure ~ border_color,
          TRUE                        ~ "grey40"
        )
      )

    # Histogram — make_cytokine_histogram() in functions_plots.R
    p_hist <- make_cytokine_histogram(d_raw, cyt, target_exposure)

    # Bar plot — make_cytokine_barplot() in functions_plots.R
    p_bar <- make_cytokine_barplot(
      summary_raw     = summary_raw,
      d_raw           = d_raw,
      cyt             = cyt,
      target_exposure = target_exposure,
      exp_short       = exp_short,
      e2_label        = e2_label,
      hormone_levels  = hormone_levels,
      border_color    = border_color,
      panel_fill_none = panel_fill_none,
      alpha_q         = ALPHA_Q,
      trend_alpha     = TREND_ALPHA,
      up_color        = UP_COLOR,
      down_color      = DOWN_COLOR,
      estradiol_fill  = ESTRADIOL_FILL
    )

    cyt_safe <- safe_name(cyt)
    cyt_subdir <- paste0("cytokine_plots_", safe_exp)
    ensure_dir(here::here(PATH_OUTPUT_FIGS, cyt_subdir))

    save_plot(file.path(cyt_subdir, paste0("histogram_", cyt_safe, ".png")),
              p_hist, width = 8, height = 5, dpi = 150, bg = "white")
    save_plot(file.path(cyt_subdir, paste0("barplot_",   cyt_safe, ".png")),
              p_bar,  width = 8, height = 6, dpi = 150, bg = "white")
  }

  cat("✓ Done:", target_exposure, "\n\n")
}
```

------------------------------------------------------------------------

# Variance Analysis

Uses `variancePartition::fitExtractVarPartModel()` to partition cytokine variance across SMOKER, SEX, HORMONE, TIMEPOINT, CELLTYPE, EXPOSURE, and PATIENTCODE. Loops over all four high-dose exposures. PCA plots use the last exposure's matrix (Red Oak 25 by default).

```{r smoker-variance, fig.width=14, fig.height=10}
stopifnot("SMOKER" %in% names(WSTC_full))

vp_results_list <- list()
vp_summary_list <- list()

for (exp in plot_exposures) {

  cat("\n── Variance partition:", exp, "──\n")

  df_exp <- imp$data %>%
    dplyr::filter(
      SEX      %in% c("F", "M"),
      EXPOSURE %in% c(PBS_LEVEL, exp)
    ) %>%
    dplyr::mutate(
      SMOKER      = factor(SMOKER),
      SEX         = factor(SEX,      levels = SEX_LEVELS),
      HORMONE     = factor(HORMONE,  levels = HORMONE_LEVELS),
      TIMEPOINT   = factor(TIMEPOINT, levels = TIMEPOINT_LEVELS),
      CELLTYPE    = factor(CELLTYPE, levels = CELLTYPE_LEVELS),
      PATIENTCODE = factor(PATIENTCODE),
      EXPOSURE    = factor(EXPOSURE)
    )

  if (!"SMOKER" %in% names(df_exp) ||
      length(unique(df_exp$SMOKER)) < 2) {
    cat("  ⚠ SMOKER not available or has < 2 levels; skipping.\n")
    next
  }

  cyt_mat <- df_exp %>%
    dplyr::select(dplyr::all_of(cytokine_cols_all)) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(),
                                ~ log2(as.numeric(.) + PSEUDOCOUNT))) %>%
    as.matrix()
  # Preserve sample IDs so downstream PCA join is key-based, not positional
  rownames(cyt_mat) <- df_exp$SAMPLEID
  cyt_mat <- t(cyt_mat)   # variancePartition wants features × samples

  rv      <- apply(cyt_mat, 1, var, na.rm = TRUE)
  cyt_mat <- cyt_mat[rv > 0 & !is.na(rv), ]
  cat("  Retained", nrow(cyt_mat), "cytokines\n")

  vp_form <- ~ (1 | SMOKER) + (1 | SEX) + (1 | HORMONE) + (1 | TIMEPOINT) +
               (1 | CELLTYPE) + (1 | EXPOSURE) +
               (1 | PATIENTCODE)

  vp_fit <- tryCatch(
    variancePartition::fitExtractVarPartModel(cyt_mat, vp_form, df_exp),
    error = function(e) {
      cat("  ⚠ Failed:", conditionMessage(e), "\n")
      NULL
    }
  )
  if (is.null(vp_fit)) next

  vp_results_list[[exp]] <- vp_fit
  vp_summary_list[[exp]] <- as.data.frame(vp_fit) %>%
    tibble::rownames_to_column("Cytokine") %>%
    dplyr::mutate(
      dplyr::across(where(is.numeric), ~ round(. * 100, 2)),
      Exposure = exp
    )

  cat("  ✓ Done\n")
}

# ── Outputs ───────────────────────────────────────────────────────────────────
if (length(vp_results_list) > 0) {

  vp_summary_all <- dplyr::bind_rows(vp_summary_list)
  save_table(vp_summary_all, "smoker_variance_all_exposures.csv")

  # Violin plots per exposure
  vp_plots <- lapply(names(vp_results_list), function(exp) {
    variancePartition::plotVarPart(
      variancePartition::sortCols(vp_results_list[[exp]])) +
      ggplot2::labs(title = exp) +
      ggplot2::theme(
        plot.title  = ggplot2::element_text(face = "bold",
                                            hjust = 0.5, size = 11),
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 9)
      )
  })

  p_vp_combined <- cowplot::plot_grid(
    plotlist   = vp_plots,
    nrow       = 2, ncol = 2,
    labels     = "AUTO", label_size = 12
  )
  print(p_vp_combined)
  save_plot("smoker_variance_partition_all_exposures.png",
            p_vp_combined, width = 14, height = 10, dpi = 300, bg = "white")

  # SMOKER summary dot plot
  p_smoker_dot <- vp_summary_all %>%
    dplyr::select(Cytokine, Exposure, SMOKER) %>%
    dplyr::mutate(
      Exposure = factor(Exposure, levels = plot_exposures),
      Cytokine = reorder(Cytokine, SMOKER, FUN = mean)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = Exposure, y = Cytokine,
                                 size = SMOKER, color = SMOKER)) +
    ggplot2::geom_point(alpha = 0.85) +
    ggplot2::scale_size_continuous(name   = "% Variance",
                                   range  = c(1, 8), limits = c(0, NA)) +
    ggplot2::scale_color_gradient(low    = "grey80",
                                  high   = UP_COLOR,
                                  name   = "% Variance") +
    ggplot2::labs(
      title = "SMOKER contribution to cytokine variance by fuel type",
      x     = NULL,
      y     = "Cytokine"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      axis.text.x      = ggplot2::element_text(angle = 30,
                                               hjust = 1, face = "bold"),
      axis.text.y      = ggplot2::element_text(face = "bold"),
      legend.position  = "right"
    )
  print(p_smoker_dot)
  save_plot("smoker_variance_summary_dotplot.png",
            p_smoker_dot, width = 8, height = 7, dpi = 300, bg = "white")

  # PCA (uses last loop's cyt_mat / df_exp)
  pca_mat <- t(cyt_mat)
  cv      <- apply(pca_mat, 2, var, na.rm = TRUE)
  pca_mat <- pca_mat[, cv > 0 & !is.na(cv)]
  pca_res <- prcomp(pca_mat, center = TRUE, scale. = TRUE)
  var_exp <- summary(pca_res)$importance[2, 1:2] * 100

  pca_df <- as.data.frame(pca_res$x[, 1:2]) %>%
    tibble::rownames_to_column("SAMPLEID") %>%
    dplyr::left_join(
      df_exp %>%
        dplyr::select(SAMPLEID, SMOKER, SEX, HORMONE, TIMEPOINT, CELLTYPE, EXPOSURE, PATIENTCODE),
      by = "SAMPLEID"
    )

  # PCA A: coloured by SMOKER
  p_pca_smoker <- ggplot2::ggplot(
    pca_df,
    ggplot2::aes(PC1, PC2, fill = SMOKER, color = SMOKER, shape = SEX)
  ) +
    ggplot2::stat_ellipse(ggplot2::aes(group = SMOKER),
                          geom = "polygon", alpha = 0.12,
                          linewidth = 0.6, show.legend = FALSE) +
    ggplot2::geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    ggplot2::scale_fill_manual(
      values = c("Y" = UP_COLOR,   "N" = DOWN_COLOR),
      labels = c("Y" = "Smoker",   "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::scale_color_manual(
      values = c("Y" = UP_COLOR,   "N" = DOWN_COLOR),
      labels = c("Y" = "Smoker",   "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::scale_shape_manual(
      values = c("M" = 21, "F" = 22),
      labels = c("M" = "Male", "F" = "Female"),
      name   = "Sex") +
    ggplot2::labs(
      x        = paste0("PC1 (", round(var_exp[1], 1), "%)"),
      y        = paste0("PC2 (", round(var_exp[2], 1), "%)"),
      title    = "PCA \u2014 Cytokine profiles by smoking status",
      subtitle = paste0("Last exposure: ", tail(names(vp_results_list), 1),
                        " + PBS")
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      plot.subtitle    = ggplot2::element_text(hjust = 0.5, size = 10,
                                               color = "grey40"),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      panel.grid.minor = ggplot2::element_blank()
    )

  # PCA B: coloured by EXPOSURE
  p_pca_exposure <- ggplot2::ggplot(
    pca_df,
    ggplot2::aes(PC1, PC2, fill = EXPOSURE, color = EXPOSURE, shape = SMOKER)
  ) +
    ggplot2::stat_ellipse(ggplot2::aes(group = EXPOSURE),
                          geom = "polygon", alpha = 0.12,
                          linewidth = 0.6, show.legend = FALSE) +
    ggplot2::geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    ggplot2::scale_fill_manual(
      values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
      name   = "Exposure") +
    ggplot2::scale_color_manual(
      values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
      name   = "Exposure") +
    ggplot2::scale_shape_manual(
      values = c("Y" = 24, "N" = 21),
      labels = c("Y" = "Smoker", "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::labs(
      x        = paste0("PC1 (", round(var_exp[1], 1), "%)"),
      y        = paste0("PC2 (", round(var_exp[2], 1), "%)"),
      title    = "PCA \u2014 Cytokine profiles by exposure",
      subtitle = "Shape = smoking status"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      plot.subtitle    = ggplot2::element_text(hjust = 0.5, size = 10,
                                               color = "grey40"),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      panel.grid.minor = ggplot2::element_blank()
    )

  p_pca_combined <- cowplot::plot_grid(
    p_pca_smoker, p_pca_exposure,
    nrow       = 1,
    labels     = c("A", "B"),
    label_size = 14
  )
  print(p_pca_combined)
  save_plot("smoker_pca_combined.png",
            p_pca_combined, width = 14, height = 6, dpi = 300, bg = "white")

  # Decision guide
  cat("\n=== SMOKER DECISION GUIDE ===\n")
  cat("SMOKER >5% variance in ≥1 cytokine\n")
  cat("  → Add SMOKER as fixed effect in screen_one_exposure_lmer_log2()\n")
  cat("  → Formula: resp ~ EXPOSURE + SMOKER + (1|PATIENTCODE)\n\n")

  cat("Cytokines where SMOKER explains >5% variance:\n")
  vp_summary_all %>%
    dplyr::filter(SMOKER > 5) %>%
    dplyr::select(Exposure, Cytokine, SMOKER) %>%
    dplyr::arrange(dplyr::desc(SMOKER)) %>%
    print()

} else {
  cat("⚠ No variance partition results to plot.\n")
}
```

------------------------------------------------------------------------

```{r pca-by-factors, fig.width=14, fig.height=10}
# ── PCA across ALL exposures/celltypes/hormones (independent of smoker loop) ──
# Uses the global `imp$data` / `cytokine_cols_all` from cytokine-setup, so it
# is NOT tied to whichever exposure happened to run last in the SMOKER loop.

pca_full_df <- imp$data %>%
  dplyr::filter(SEX %in% c("F", "M")) %>%
  dplyr::mutate(
    SEX         = factor(SEX,      levels = SEX_LEVELS),
    HORMONE     = factor(HORMONE,  levels = HORMONE_LEVELS),
    TIMEPOINT   = factor(TIMEPOINT, levels = TIMEPOINT_LEVELS),
    CELLTYPE    = factor(CELLTYPE, levels = CELLTYPE_LEVELS),
    EXPOSURE    = factor(EXPOSURE),
    PATIENTCODE = factor(PATIENTCODE)
  )

cyt_mat_full <- pca_full_df %>%
  dplyr::select(dplyr::all_of(cytokine_cols_all)) %>%
  dplyr::mutate(dplyr::across(dplyr::everything(),
                              ~ log2(as.numeric(.) + PSEUDOCOUNT))) %>%
  as.matrix()
rownames(cyt_mat_full) <- pca_full_df$SAMPLEID

cv_full      <- apply(cyt_mat_full, 2, var, na.rm = TRUE)
cyt_mat_full <- cyt_mat_full[, cv_full > 0 & !is.na(cv_full)]

pca_res_full <- prcomp(cyt_mat_full, center = TRUE, scale. = TRUE)
var_exp_full <- summary(pca_res_full)$importance[2, 1:2] * 100

pca_df_full <- as.data.frame(pca_res_full$x[, 1:2]) %>%
  tibble::rownames_to_column("SAMPLEID") %>%
  dplyr::left_join(
    pca_full_df %>%
      dplyr::select(SAMPLEID, SEX, HORMONE, TIMEPOINT, CELLTYPE, EXPOSURE, PATIENTCODE),
    by = "SAMPLEID"
  )

pc_labs <- c(
  x = paste0("PC1 (", round(var_exp_full[1], 1), "%)"),
  y = paste0("PC2 (", round(var_exp_full[2], 1), "%)")
)

base_pca_theme <- ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold", hjust = 0.5, size = 13),
    plot.subtitle    = ggplot2::element_text(hjust = 0.5, size = 10, color = "grey40"),
    panel.border     = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.5),
    panel.grid.major = ggplot2::element_line(color = "gray88"),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text       = ggplot2::element_text(face = "bold")
  )

# ── A) By EXPOSURE (wood fuel) ────────────────────────────────────────────
p_pca_exposure_all <- ggplot2::ggplot(
  pca_df_full, ggplot2::aes(PC1, PC2, fill = EXPOSURE, color = EXPOSURE)
) +
  ggplot2::stat_ellipse(ggplot2::aes(group = EXPOSURE),
                        geom = "polygon", alpha = 0.10,
                        linewidth = 0.6, show.legend = FALSE) +
  ggplot2::geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  ggplot2::scale_fill_manual(values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
                             name = "Exposure") +
  ggplot2::scale_color_manual(values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
                              name = "Exposure") +
  ggplot2::labs(x = pc_labs["x"], y = pc_labs["y"],
                title = "PCA \u2014 By Exposure (wood fuel)") +
  base_pca_theme

# ── B) By AIRWAY (CELLTYPE) ────────────────────────────────────────────────
p_pca_celltype <- ggplot2::ggplot(
  pca_df_full, ggplot2::aes(PC1, PC2, fill = CELLTYPE, color = CELLTYPE)
) +
  ggplot2::stat_ellipse(ggplot2::aes(group = CELLTYPE),
                        geom = "polygon", alpha = 0.10,
                        linewidth = 0.6, show.legend = FALSE) +
  ggplot2::geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  ggplot2::scale_fill_manual(values = c("SAE" = DOWN_COLOR, "LAE" = UP_COLOR), name = "Airway") +
  ggplot2::scale_color_manual(values = c("SAE" = DOWN_COLOR, "LAE" = UP_COLOR), name = "Airway") +
  ggplot2::labs(x = pc_labs["x"], y = pc_labs["y"],
                title = "PCA \u2014 By Airway (SAE vs LAE)") +
  base_pca_theme

# ── C) By SEX ───────────────────────────────────────────────────────────────
p_pca_sex <- ggplot2::ggplot(
  pca_df_full, ggplot2::aes(PC1, PC2, fill = SEX, color = SEX)
) +
  ggplot2::stat_ellipse(ggplot2::aes(group = SEX),
                        geom = "polygon", alpha = 0.10,
                        linewidth = 0.6, show.legend = FALSE) +
  ggplot2::geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  ggplot2::scale_fill_manual(values = c("F" = "#D7191C", "M" = "#2C7BB6"), name = "Sex") +
  ggplot2::scale_color_manual(values = c("F" = "#D7191C", "M" = "#2C7BB6"), name = "Sex") +
  ggplot2::labs(x = pc_labs["x"], y = pc_labs["y"], title = "PCA \u2014 By Sex") +
  base_pca_theme

# ── D) By HORMONE ─────────────────────────────────────────────────────────
p_pca_hormone <- ggplot2::ggplot(
  pca_df_full, ggplot2::aes(PC1, PC2, fill = HORMONE, color = HORMONE)
) +
  ggplot2::stat_ellipse(ggplot2::aes(group = HORMONE),
                        geom = "polygon", alpha = 0.10,
                        linewidth = 0.6, show.legend = FALSE) +
  ggplot2::geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  ggplot2::scale_fill_manual(values = c("NONE" = "grey50", "Estradiol" = "#4DAF4A"), name = "Hormone") +
  ggplot2::scale_color_manual(values = c("NONE" = "grey50", "Estradiol" = "#4DAF4A"), name = "Hormone") +
  ggplot2::labs(x = pc_labs["x"], y = pc_labs["y"], title = "PCA \u2014 By Hormone") +
  base_pca_theme

# ── E) By TIMEPOINT ───────────────────────────────────────────────────────
p_pca_timepoint <- ggplot2::ggplot(
  pca_df_full, ggplot2::aes(PC1, PC2, fill = TIMEPOINT, color = TIMEPOINT)
) +
  ggplot2::stat_ellipse(ggplot2::aes(group = TIMEPOINT),
                        geom = "polygon", alpha = 0.10,
                        linewidth = 0.6, show.legend = FALSE) +
  ggplot2::geom_point(shape = 21, size = 2.5, alpha = 0.85, stroke = 0.6) +
  ggplot2::scale_fill_manual(values = c("4" = "#4DAF4A", "144" = "#984EA3"), name = "Timepoint") +
  ggplot2::scale_color_manual(values = c("4" = "#4DAF4A", "144" = "#984EA3"), name = "Timepoint") +
  ggplot2::labs(x = pc_labs["x"], y = pc_labs["y"], title = "PCA \u2014 By Timepoint") +
  base_pca_theme

p_pca_factors_combined <- cowplot::plot_grid(
  p_pca_exposure_all, p_pca_celltype, p_pca_sex, p_pca_hormone, p_pca_timepoint,
  nrow = 2, ncol = 3,
  labels = "AUTO", label_size = 14
)
print(p_pca_factors_combined)

save_plot("pca_by_exposure_airway_sex_hormone_timepoint.png",
          p_pca_factors_combined, width = 14, height = 10, dpi = 300, bg = "white")

# Individual saves in case you want them separately
save_plot("pca_by_exposure.png", p_pca_exposure_all, width = 7, height = 6, dpi = 300, bg = "white")
save_plot("pca_by_airway.png",   p_pca_celltype,      width = 7, height = 6, dpi = 300, bg = "white")
save_plot("pca_by_sex.png",      p_pca_sex,           width = 7, height = 6, dpi = 300, bg = "white")
save_plot("pca_by_hormone.png",  p_pca_hormone,       width = 7, height = 6, dpi = 300, bg = "white")
save_plot("pca_by_timepoint.png", p_pca_timepoint,    width = 7, height = 6, dpi = 300, bg = "white")
```

# Session Information

```{r session-info}
cat("════════════════════════════════════════════════\n")
cat("  01_WSTC_ANALYSIS COMPLETE\n")
cat("════════════════════════════════════════════════\n\n")
cat("Outputs saved to:\n")
cat("  Figures :", here::here(PATH_OUTPUT_FIGS),   "\n")
cat("  Tables  :", here::here(PATH_OUTPUT_TABLES), "\n")
cat("  Reports :", here::here(PATH_OUTPUT_REPORT), "\n\n")
sessionInfo()
```
